---
title: "project1"
output: html_document
date: "2024-05-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages
```{r}
library(tidyverse)
library(pracma)
library(lme4)
library(targets)
library(here)
library(data.table)
library(ggplot2)
```
	
# Code from Anders
## Load the data

```{r}
# tar_load(c(observed_sample_data, sample_data), store = here("project_1/_targets"))
```


```{r}
sample_data[, source := "Sample"]
observed_sample_data[, source := "Observed"]


combined_data <- rbind(sample_data[, -3], observed_sample_data[,-4])

ggplot(combined_data, aes(x = age, y = bmi, color = source, group = source)) +
  geom_line(size = 1.2) +  
  labs(x = "Age", y = "Average BMI", title = "BMI Trends by Age in Sample and Observed Data") +
  theme_bw() +
  theme(legend.position = "top")
```


## calculate the summary statistics
```{r}
# For the sample_data
res_calculate_summary_statistics <- calculate_summary_statistics(data = sample_data)

# For the observed_sample_data
res_calculate_summary_statistics_observed <- calculate_summary_statistics(data = observed_sample_data)
```


## Compare the summary statistics
```{r}
# Merge the two datasets by "id"
comparison <- merge(
  res_calculate_summary_statistics, 
  res_calculate_summary_statistics_observed, 
  by = "id", 
  suffixes = c("_sample", "_observed")
)


metrics <- colnames(res_calculate_summary_statistics)[-c(1:2)]

RMSE_results_by_group <- list()
overall_RMSE_results <- list()

for (metric in metrics) {
  sample_col <- paste0(metric, "_sample")
  observed_col <- paste0(metric, "_observed")
  squared_diff_col <- paste0(metric, "_squared_diff")
  RMSE_col <- paste0(metric, "_RMSE")  
  
  comparison[[squared_diff_col]] <- (comparison[[sample_col]] - comparison[[observed_col]])^2
  
  RMSE_by_group <- comparison[, 
                              .(Root_Mean_Squared_Error = sqrt(mean(get(squared_diff_col), na.rm = TRUE)), 
                                group = count_observed), 
                              by = count_observed]
  
  RMSE_results_by_group[[RMSE_col]] <- RMSE_by_group$Root_Mean_Squared_Error
  
  overall_RMSE_value <- sqrt(mean(comparison[[squared_diff_col]], na.rm = TRUE))
  overall_RMSE_results[[RMSE_col]] <- overall_RMSE_value
}


RMSE_results_by_group_df <- as.data.frame(RMSE_results_by_group)
RMSE_results_by_group_df$group <- unique(comparison$count_observed)

overall_RMSE_results_df <- as.data.frame(overall_RMSE_results)
overall_RMSE_results_df$group <- "Overall"


RMSE_results <- 
  rbind(RMSE_results_by_group_df, overall_RMSE_results_df) %>%
  mutate(group = factor(group, levels = c(1:7, "Overall"))) %>%
  arrange(group)
RMSE_results
```


```{r}
RMSE_long <- melt(RMSE_results, 
                  id.vars = "group", 
                  variable.name = "Variable", 
                  value.name = "RMSE") 

RMSE_filtered <- subset(RMSE_long, group != 'Overall' & group != 1)

ggplot(subset(RMSE_long, group != '1'), aes(x = group, y = RMSE)) +
  geom_point(aes(color = ifelse(group == "Overall", "Overall", "Other")), size = 3, show.legend = FALSE) +  
  geom_line(data = RMSE_filtered, aes(group = Variable), size = 1) + 
  labs(x = "Group", y = "RMSE", title = "RMSE of Each Variable") +
  facet_wrap(~Variable, scales = "free_y") + 
  scale_color_manual(values = c("Overall" = "red", "Other" = "black")) +
  theme_bw()

ggsave("RMSE.pdf", width = 15, height = 10, dpi = 300)
```


```{r}
summary_stats <- function(x) {
  c(
    Mean = mean(x, na.rm = TRUE),
    SD = sd(x, na.rm = TRUE),
    Q25 = quantile(x, 0.25, na.rm = TRUE), 
    Median = median(x, na.rm = TRUE),
    Q75 = quantile(x, 0.75, na.rm = TRUE),  
    Min = min(x, na.rm = TRUE),
    Max = max(x, na.rm = TRUE)
  )
}

summary_results <- sapply(c(paste0(metrics, '_sample'), paste0(metrics, '_observed')), 
                          function(col) summary_stats(comparison[[col]]))%>% as.data.frame()

summary_results
```


## Box plot
```{r}
# Melt the comparison dataset to long format
comparison_melted <- melt(comparison, 
                          id.vars = "id", 
                          measure.vars = c(paste0(metrics, '_sample'),paste0(metrics, '_observed')), 
                          variable.name = "metric_type", 
                          value.name = "value")[value != 0]

setDT(comparison_melted)
comparison_melted[, source := ifelse(grepl("_sample$", metric_type), "Sample", "Observed")]
comparison_melted[, metric := gsub("_sample|_observed", "", metric_type)]

# Plot using ggplot2
ggplot(comparison_melted, aes(y = value, fill = source, x = source)) +
  geom_boxplot() + 
  facet_wrap(~metric, scales = "free_y") +  # Free scale on Y axis for each plot
  theme_minimal() +
  labs(x = "source", y = "value", fill = "source", title = "Comparison between the sample data and observed data") +
  scale_fill_manual(values = c("Sample" = "blue", "Observed" = "red")) +
  theme_bw()+
  theme(legend.position = "top")

ggsave("box_plots_nozero.pdf", width = 12, height = 12, dpi = 300)
```



We first use the true curves to calculate the true AUC for each individual using
the trapezoidal rule:


## Fit a linear mixed model

We fit a simple linear mixed model to the data. Here I use the same model as you
specify below.

```{r}
lmm1 <- lmer(bmi ~ age + I(age^2) + (age | id), data = observed_sample_data)
```

The model warns that it failed to converge, but I don't know if this is
problematic. When looking at the results below it does not seem like anything
went wrong.

We can visualize the results from the model and compare to the observed data as
follows. The black dots are the observed data, the blue lines are the inter- and
extrapolation made by the linear mixed model.

```{r}
pred_grid1 <- expand.grid(id = unique(observed_sample_data[, id]),
                         age = seq(40, 60, length.out = 100))
setDT(pred_grid1)
pred_grid1[, bmi := predict(lmm1, newdata = .SD)]
plot_lmm1 <- ggplot(data = observed_sample_data, aes(x = age, y = bmi)) +
    theme_bw() +
    facet_wrap(~id) +
    geom_point() +
    geom_line(data = pred_grid1, color = "blue")
plot_lmm1
```

We can also compare the fitted model to the true curves.

```{r}
plot_lmm1 + geom_line(data = sample_data)
```

## Estimate the AUC based on the fitted linear mixed model
Using the predicted BMI values from the linear mixed model we can obtain the AUC
for each individual according to the fitted linear mixed model:

```{r}
est_AUC1 <- pred_grid1[, .(AUC = calc_AUC(age, bmi)), by = id]
est_AUC1
```

We can compare the estimated individual AUC's to the true individual AUC's. The
black dots are the true individual AUC's and the blue dots are the estimated
ones. The red curve illustrates the estimation error.

```{r}
AUC_error1 <- merge(true_AUC[, .(id, true_AUC = AUC)], est_AUC1[, .(id, est_AUC = AUC)], by = "id")
AUC_error1[, error := true_AUC-est_AUC]
ggplot(AUC_error1, aes(x = id)) +
    theme_bw() +
    geom_segment(aes(xend = id, y = true_AUC, yend = est_AUC), col = "red") + 
    geom_point(aes(y = true_AUC), col = "black") +
    geom_point(aes(y = est_AUC), col = "blue")  +
    ylab("AUC")
```

We can also look at the bias and the variance of the estimator:

```{r}
mean(true_AUC[, AUC] -est_AUC1[, AUC]) # Bias
var(est_AUC1[, AUC]) # standard deviation 
```

But I think we need to think a bit about how we compare performance, because
this is looking at the bias and variance across individuals.

# Calculate AUC

```{r}
fun_auc <- function(data) {
  
  data %>%
    arrange(id, age) %>%
    group_by(id) %>%
    mutate(next_age = lead(age),
           next_bmi = lead(bmi),
           auc = if_else(!is.na(next_age), 0.5 * (bmi + next_bmi) * (next_age - age), NA_real_)) %>%
    ungroup()
}

observed_sample_data <- 
  fun_auc(observed_sample_data) %>% 
  rename('observed_auc' = 'auc')

sample_data <- 
  fun_auc(sample_data) %>% 
  rename('true_auc' = 'auc') %>% 
  merge(., observed_sample_data[,c('id', 'age', 'observed_auc')], by = c('id', 'age'), all = TRUE)
```


# Prediction

## Linear mixed model
```{r}
# linear mixed model
lmm <- lmer(observed_auc ~ age + I(age^2) + (1 + age | id), data = sample_data, REML = FALSE)

# predicting auc and sum up per individual
auc <- 
  sample_data %>% 
  subset(age != 60) %>% # no next_bmi to compute true_auc
  mutate(predicted_auc = predict(lmm, newdata = sample_data, re.form = ~ (1 + age_centered | id))) %>%
  group_by(id) %>%
  summarise(true_auc = sum(true_auc, na.rm = T),
           predicted_auc = sum(predicted_auc))
```



```{r}
# Calculate Mean Squared Error (MSE)
auc <- auc %>% mutate(MSE = (predicted_auc - true_auc)^2)

boxplot(auc$MSE, 
        main = "MSE of Predicted AUC vs. True AUC",
        ylab = "MSE")

cat("Median MSE:", median(auc$MSE), "\n")
```



